Tidy data structure to support exploration and modeling of temporal data

### 2018-12-23 main.Rmd
updated flowcharts

Suppose that temporal data consists of $N$ subjects/observational units. Given subject $i$ $(i = 1, 2, \dots, N)$, $X_{ijt}$ denotes the $j$th measurement $(j = 1, 2, \dots, p)$ indexed at a unique sequence of time points $0 \le t \le T_{i}$. Temporal data arrives in many possible formats, so does time. For example, time can carry with various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains rich information: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors. @r4ds succinctly describes the data science workflow given in Figure \ref{fig:pipeline-ann}. It is annotated to show that the divide between data that feeds into the beginning of analysis and the current time series objects that fits into modelling is significant, making data preparation for modeling and forecasting painful.

```{r pipeline-ann, fig.cap = "(ref:pipeline-ann-cap)"}
include_graphics("img/time-series-structure.pdf")
```

(ref:pipeline-ann-cap) An overview of data science program. The existing standards are model-focused, resulting misalliance between data and model.

The existing standards, particularly time series objects, are model-oriented with little support for non-numeric values, leaning its data abstraction towards matrix. Time series have been isolated from general data problems due to this data choice. Analysts who start analysing time series would be pushed to learn new transformation and visualization toolkits that are different from what they used to operate.

@@

The **tidyverse** way of thinking largely formalizes exploratory data analysis. Tidy concepts have been extended to other domains in R community.

Temporal data is integrated to tidy data and the grammar. @r4ds argues that 80% of data analysis tasks can be solved with tidy tools while the remaining 20% requires other tools.

@@

The choice of tidy representation of temporal data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. Contextual semantics are introduced to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series model inputs. Index, key and time interval are the three stone pillars to this new semantically-structured temporal data, which we name as "tsibble", inspired by "ts" and "tibble".

@@

A set of verbs have been implemented to conduct temporal operations. There are three kinds of sliding window operations:

@@

Time series are represented as matrices, with standards being provided by native `ts` object in R, extended by **zoo** [@zoo2005], and **xts** [@R-xts]. As discussed in Section \ref{sec:time-series}, this data organization sets far apart from data origin. Due to time isolated from the main data, the supporting functions that intend to deal with time must be applied to whole data. That data-centric & human-centered.

>>>>

Suppose that temporal data consists of $N$ subjects/observational units. Given subject $i$ $(i = 1, 2, \dots, N)$, $X_{ijt}$ denotes the $j$th measurement $(j = 1, 2, \dots, p)$ indexed at a unique sequence of time points $0 \le t \le T_{i}$. Temporal data arrives in many possible formats, so does time. For example, time can carry with various time resolutions (hours, minutes, and seconds), and they are typically associated with different time zones with adjustments like summer time. It could be irregularly recorded, which is particularly true with longitudinal measurements like patient visits to a doctor's office. Temporal data also often contains rich information: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors. @r4ds succinctly describes the data science workflow given in Figure \ref{fig:pipeline-ann}. It is annotated to emphasize the significant divide between data that feeds into the beginning of analysis and the current time series objects that fits into modelling, making data preparation for modeling and forecasting painful.

```{r pipeline-ann, fig.cap = "(ref:pipeline-ann-cap)"}
include_graphics("img/motivation.pdf")
```

(ref:pipeline-ann-cap) An overview of the data science program centering around data. Temporal data should feed into the import stage, but the existing time series objects are model-focused, thus resulting misalliance between data and model.

Current time series objects, which are model-oriented matrices with little support for non-numeric values, feed into the framework solely at the modelling stage. Analysts are expected to do their own data pre-processing and take care of anything else needed to get to this stage, which leads to a myriad ad-hoc solutions and duplicated efforts. The diagram correspondingly would place the model at the centre of the analytical universe, and all the transformations, visualisations would hinge on that format.

@@

The **tidyverse** largely formalizes exploratory data analysis. The whole R community has quickly adopted the **tidyverse** way of thinking to much broader domains, such as simple features for spatial data in the **sf** package [@RJ-2018-009] and missing value handling in the **naniar** package [@tierney-naniar-2018]. Temporal data is left behind in this regard. This paper proposes a new tidy representation for temporal data and a unified framework to streamline the workflow from data pre-processing to visualization and forecasting, as an integral part of tidy data analysis.

@@

```{r tsibble-fit, fig.cap = "(ref:tsibble-fit-cap)"}
include_graphics("img/tsibble-fit.pdf")
```

(ref:tsibble-fit-cap) Tsibble's abstraction arises from a data-centric point of view, which could neatly embed into the tidyverse ecosystem. It should be noted that tsibble positions itself at the "tidy" step rather than "import", which ensures valid tidy data to time series analytics and models.

The choice of tidy representation of temporal data arises from a data-centric perspective, which accommodates all of the operations that are to be performed on the data. Contextual semantics are introduced to tidy data in order to support more intuitive time-related manipulations and enlighten new perspectives for time series model inputs. Index, key and time interval are the three stone pillars to this new semantically-structured temporal data, which we name as "tsibble", inspired by "ts" and "tibble".

@@

Temporal data often involves moving window calculations. A set of verbs have been implemented to allow for different variations of sliding windows operations:

@@

Matrices are a conventional way of organizing time series, with standards being provided by native `ts` object in R, extended by **zoo** [@zoo2005], and **xts** [@R-xts]. Time indices are stored as an attribute, separating from the main body of the data.

data-centric & human-centered.
